CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems

### 2014-11-18 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
added all architecture experiment results
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4507 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\section{Architecture models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are left untouched, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

@@

%   - Interposer-Routing interface
\subsection{Interposer-Routing interface}
In a simple architecture, the vertical routing wires that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

\begin{enumerate}[label=(\alph*)]
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. Hence, when this variable is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect some adjacent ``dangling" vertical routing wires to the multiplexer. This allows for additional flexibility in routing. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 multiplexer at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

\item Fan-out Transfer for Interposer Wires\\
When this option is \textit{on}, we allow the interposer wire output to drive not only its own corresponding vertical wire at the other side of the cut, but also the other ``dangling" vertical routing wires that are not driven by anything because their corresponding interposer wires are cut.

@@

\section{CAD Enhancements}
\label{cadSection}

\subsection{Placer Optimization}
\label{cad_enh_placer_subsection}
Placement is crucial in mitigating the impact of the reduced wiring and increased delay when crossing between dice; a good placement should minimize the number of signals crossing between dice, particularly time-critical ones. VPR uses 2 different costs as the metrics for its placer algorithm: the timing cost and the bounding box (wiring) cost. We modified the placer's bounding box and timing costs to account for the increased latency to cross the interposer and for the reduced wire capacity close to the cutlines, as the wire availability becomes more sparse.

@@

\subsection{Router Optimization}
Once the Routing Resource Graph is modified as detailed in Section \ref{archSection}, the VPR router adapted automatically to the interposer architecture. The router starts at a net source node and expands by discovering adjacent routing nodes until it reaches the destination node. At any point, the router knows the cost of the path taken to the current point and it uses a \textit{lookahead cost function} to estimate the resource and delay cost from the current node to the destination. Using this lookahead cost, the router can properly sort routing alternatives. The lookahead cost function in VPR is not aware of the interposer, and therefore it takes a long time to converge to a routing decision. To make the lookahead aware of the interposer, we modify the lookahead cost function by adding terms that account for extra \textit{interposer delay} and extra \textit{interposer nodes} that may exist on the route from the current node to the destination node.

@@

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the "flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}). The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

We used VPR 7.0, with the enhancements we detailed above, and the architecture file \textit{k6\_frac\_N10\_mem32K\_40nm} in the experiments below. All experiments targeted the smallest FPGA (with number of rows equal to number of columns) which could accomodate a benchmark circuit; this represents a very full FPGA with little white space left, and hence presents a difficult case to an interposer-based FPGA as no die can be left mostly empty.

Figures \ref{fig:minW_constant_sweep} and \ref{fig:areadelay_constant_sweep} show the efficacy of the five placer wiring cost modifications as their weight, \textit{C} varies. Each point in the graphs is the geometric mean of the results with \textit{\% wires cut} = 60 and 80, and the results for 60\% and 80\% are themselves geometric means of the results of 6 circuits from the VTR benchmark suite~\cite{vtr2012}:  \textit{stereovision0}, \textit{stereovision1}, \textit{mkDelayWorker32B}, \textit{mkSMAdapter4B}, \textit{or1200} and \textit{blob\_merge}. These values for \textit{\% wires cut} were chosen because they in the likely range of wires which can cross the interposer without exceeding the microbump capacity, as described in Section~\ref{virtex7section}. The \textit{area-delay product} is the product of the \textit{minimum channel width} and the \textit{critical path delay}. To obtain the \textit{critical path delay} the circuits were run with a low stress routing with a channel width, $W = 1.3 \times minW$, where $minW$ is the minimum channel width for which the circuit is still routable.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{minW_constant_sweep_log.eps}
\caption{Minimum channel width vs. weighting for different \textit{cut\_cost} terms.}
\label{fig:minW_constant_sweep}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{areadelay_constant_sweep_log.eps}
\caption{Area-delay product vs. weighting for different \textit{cut\_cost} terms.}
\label{fig:areadelay_constant_sweep}
\end{figure}

\begin{table}[htbp]
\begin{tabular}{|r|r|r|r|r|}
\hline
\multicolumn{1}{|l|}{Term} & \multicolumn{1}{l|}{Best C} & \multicolumn{1}{l|}{minW} & \multicolumn{1}{l|}{crit\_path(ns)} & \multicolumn{1}{l|}{Area-delay} \\ \hline \hline
None & - & 124.636 & 9.844 & 1226.955 \\ \hline
\ref{eq:cost0} & 0.3 & 122.929 & 9.413 & 1157.09 \\ \hline
\ref{eq:cost1} & 5 & 114.444 & 9.565 & 1094.68 \\ \hline
\ref{eq:cost2} & 1 & 107.303 & 9.285 & 996.26 \\ \hline
\ref{eq:cost4} & 1 & 113.381 & 9.280 & 1052.16 \\ \hline
\ref{eq:cost5} & 5 & 112.703 & 9.841 & 1109.11 \\ \hline
\end{tabular}
\caption{Best weighting and performance for each \textit{cut\_cost} term.}
\label{table:constant_sweep}
\end{table}

The best \textit{C} value and the resulting performance for each term is summarized in Table~\ref{table:constant_sweep}. Note that cost term (\ref{eq:cost2}) with weighting $C = 1.0$ has the best performance; this is the configuration used in the rest of our experiments.

We believe the key to the good performance of this cost function is that it produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline and hence are penalized equally by (\ref{eq:cost0}), while bounding box (c) does not cross a cutline and hence is not penalized. However, bounding box (b) is mostly on the lower side of the cutline; it is more likely that later smaller placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function terms (\ref{eq:cost2}) and (\ref{eq:cost4}) will both penalize bounding box (a) more than bounding box (b) to guide placement to gradually move bounding boxes to one side or the other of a cutline. Figure \ref{fig:areadelay_constant_sweep} shows that these are the two best performing \textit{cut\_cost} terms.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{bb.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

Figure \ref{fig:comparison_minW} shows the minimum channel width for different values of \textit{\% wires cut} when there are no optimizations and when \textit{cut\_const} (\ref{eq:cost2}) is used. % comment about impact vs % wires cut
Figure \ref{fig:comparison_bars} compares result quality with both enhanced timing and wiring cost functions. These experiments were run with term (\ref{eq:cost2}), and are the geometric means of the runs with \textit{\% wires cut = 60, 70, 80}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{comparison_bars_relative.eps}
\caption{Performance of different placer optimizations, relative to the original VPR results. The area-delay product is calculated as the product $minW \times critical path delay.$}
\label{fig:comparison_bars}
\end{figure}

The graphs show that combining the timing with the bounding box optimizations led to an average of 20\% improvement on the area-delay product and the most significant contribution was made by incorporation of \textit{cut\_cost} into the wire cost function.

%----------------------------------------------------
\section{Architecture results}
\label{resultsSection}

Using the best CAD settings found in Section~\ref{cadSection} we analyzed the impact of three key architecture parameters:  \textit{\% wires cut}, \textit{delay increase} and \textit{number of cuts}. All the experiments were run with the eight largest circuits from the VTR benchmark~\cite{vtr2012}, namely: \textit{bgm}, \textit{LU8PEEng}, \textit{LU32PEEng}, \textit{mcml}, \textit{mkDelayWorker32B}, \textit{stereovision0}, \textit{stereovision1} and \textit{stereovision2}. The size of these circuits ranges from $9,100$ to $153,000$ primitives (LUTs, FFs, etc.), with an average size of $52,600$ primitives. All results are the geometric mean over all circuits for a given interposer architecture.

To analyze the impact of the number of cut wires we ran experiments varying only this parameter while leaving the number of cuts and the increased delay constant. The values used for these parameters were \textit{number of cuts} $= 3$ (corresponding to four dice) and \textit{increased delay} $= 1ns$, as these values are similar to those of the \textit{XC7V2000T} device.

Figure \ref{fig:standard_minW} shows the graph of minimum channel width versus \textit{\% wires cut}. It can be noted from this graph that the minimum channel width increases slowly up to 60\% of wires cut, indicating that the placement engine is able to avoid saturating the interposer routing until that point. When more than 60\% of the wires are cut however (i.e. the interposer provides less than 40\% of the usual within-die routing capacity), the minimum channel width grows rapidly indicating that the interposer routing bandwidth has become a limiting factor. The minW value at 60\% is 20\% greater than with no wires cut, while at 70\% it is 52\% above and at 80\% it is 125\% larger than the minW with no cut wires.

The critical path delay, depicted in Figure~\ref{fig:standard_crit}, on the other hand, is not strongly influenced by the percentage of wires cut, as the critical path delay at 60\% of the wires cut is essentially the same as at 0\% wires cut. At 80\% of the wires cut the critical path delay does rise somewhat, by 6\%; this is due to two factors: the placement is being modified to improve routability at the expense of timing, and some circuitous routes are occuring due to saturated interconnect across the interposer.
%Andre, what is the critical path delay with no interposer (i.e. 0% cut and 0 delay)?

Tables \ref{table:standard_minW} and \ref{table:standard_path} show the minimum channel width and critical path delay for each circuit. The trends for individual circuits follow those of the averages quite closely.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{standard_minW.eps}
\caption{Minimum channel width vs. \textit{\% wires cut} for 3 cuts and $1ns$ of \textit{delay increase}.}
\label{fig:standard_minW}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{standard_crit_path.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 3 cuts and $1ns$ of \textit{delay increase}.}
\label{fig:standard_crit}
\end{figure}

\begin{table}[!htbp]
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\% wires cut & 0 & 20 & 40 & 60 & 70 & 80 \\ \hline \hline
bgm & 114 & 128 & 124 & 132 & 152 & 226 \\ \hline
LU32PEEng & 172 & 178 & 182 & 218 & 304 & 426 \\ \hline
LU8PEEng & 110 & 116 & 118 & 128 & 158 & 236 \\ \hline
mcml & 108 & 124 & 128 & 116 & 152 & 218 \\ \hline
mkDelayWorker32B & 76 & 80 & 82 & 88 & 100 & 132 \\ \hline
stereovision0 & 62 & 60 & 70 & 88 & 110 & 166 \\ \hline
stereovision1 & 104 & 108 & 116 & 120 & 178 & 312 \\ \hline
stereovision2 & 162 & 164 & 168 & 194 & 244 & 360 \\ \hline
Geometric mean & 108 & 114 & 118 & 129 & 164 & 243 \\ \hline
\end{tabular}
\caption{Minimum channel width when \textit{number of cuts = 3} and \textit{increased delay = 1ns}.}
\label{table:standard_minW}
\end{table}

\begin{table}[!htbp]
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\% wires cut & 0 & 20 & 40 & 60 & 70 & 80 \\ \hline \hline
bgm & 35.1 & 33.6 & 32.2 & 32.9 & 37.8 & 40.3 \\ \hline
LU32PEEng & 115 & 114 & 114 & 114 & 115 & 113 \\ \hline
LU8PEEng & 118 & 117 & 119 & 116 & 117 & 124 \\ \hline
mcml & 82.3 & 82.9 & 84.0 & 83.4 & 88.3 & 85.7 \\ \hline
mkDelayWorker32B & 11.0 & 12.3 & 12.0 & 12.3 & 14.4 & 12.8 \\ \hline
stereovision0 & 6.9 & 7.8 & 7.0 & 7.4 & 7.6 & 7.0 \\ \hline
stereovision1 & 11.6 & 9.2 & 10.4 & 11.0 & 11.6 & 11.8 \\ \hline
stereovision2 & 23.3 & 25.4 & 23.9 & 24.9 & 25.6 & 24.8 \\ \hline
Geometric mean & 30.8 & 30.9 & 30.6 & 31.2 & 33.1 & 32.6 \\ \hline
\end{tabular}
\caption{Critical path delay (in \textit{ns}) when \textit{number of cuts = 3} and \textit{increased delay = 1ns}.}
\label{table:standard_path}
\end{table}

Figure \ref{fig:crossingwires} provides an alternative way to visualize the relationship between interposer routing supply and routability. This figure shows how the geometric average minimum channel width required within the FPGA dice varies as the geometric average of the \emph{absolute} number of wires crossing the interposer in each channel varies, again for 4-die system. When 108 tracks cross the interposer, the interposer channels have the same capacity as the vertical routing channels within each FPGA die. As fewer wires cross the interposer, the channel width required within the FPGA dice increases to compensate for the routing difficulty in crossing the interposer. The increase is gradual as the interposer routing is reduced from 108 tracks per channel to 52 tracks per channel; over this range the routing per channel required in the FPGA dice increases from 108 tracks per channel to 129 tracks per channel. As the routing crossing the interposer is further reduced however, it becomes very difficult to increase the within-die routing sufficiently to compensate. At 49 tracks crossing the interposer channels, for example, the within-die routing must have a channel width of \emph{243} tracks to successfully route the designs. Clearly the CAD tools have the ability to trade-off interposer routing for within-die routing over a reasonable range but below a certain level (48\% of the original within-die minimum channel width in our experiments) routability becomes almost solely limited by the wiring crossing the interposer and further reduction in interposer routing is not productive.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{numberofcrossingwires.eps}
\caption{Geometric mean of required intra-die minimum channel width vs. geometric mean of the number of wires crossing the interposer for 3 cuts and $1ns$ of \textit{delay increase}.}
\label{fig:crossingwires}
\end{figure}

To investigate the impact of the interposer delay (\textit{delay increase}), we kept the \textit{number of cuts} constant at $3$ while \textit{delay increase} was swept between $0$ and $1.5ns$.

Figure \ref{fig:delays_crit} shows critical path delay versus \textit{\% wires cut} for 4 different values of \textit{delay increase}. The penalty in critical path delay is significant, ranging between $5$ and $9$ times the interposer \textit{delay increase}, when compared to the case where the interposer adds no delay. Note that the 0\% wires cut with a 0 ns \textit{delay increase} in Figure \ref{fig:delays_crit} corresponds to a traditional monolithic FPGA. The speed of an interposer-based FPGA is strongly correlated to \textit{delay increase}: a $0.5ns$ interposer delay increases the critical path delay by 20\%, while a $1ns$ interposer delay increases critical path delay by approximately 35\% vs. a monolithic FPGA. Once again, the critical path increase shows little correlation to the \textit{\% wires cut}, however. Tables \ref{table:delay0}, \ref{table:delay500}, \ref{table:standard_path} and \ref{table:delay1500} show the critical path delay for each circuit for each of the different values of \textit{delay increase}. Once again the results for individual circuits closely parallel the overall averages.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{delays_crit_path.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 3 cuts and $0.0$, $0.5$, $1.0$ and $1.5ns$ of \textit{delay increase}.}
\label{fig:delays_crit}
\end{figure}

\begin{table}[!htbp]
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\% wires cut & 0 & 20 & 40 & 60 & 70 & 80 \\ \hline  \hline
bgm & 25.9 & 26.3 & 25.8 & 26.1 & 25.8 & 26.7 \\ \hline
LU32PEEng & 113 & 113 & 117 & 118 & 113 & 114 \\ \hline
LU8PEEng & 115 & 114 & 116 & 115 & 118 & 115 \\ \hline
mcml & 77.2 & 78.7 & 79.5 & 80.5 & 77.3 & 76.7 \\ \hline
mkDelayWorker32B & 7.8 & 7.6 & 7.5 & 7.7 & 7.5 & 7.6 \\ \hline
stereovision0 & 4.5 & 4.3 & 4.5 & 4.5 & 4.5 & 4.5 \\ \hline
stereovision1 & 6.0 & 6.2 & 5.9 & 6.3 & 6.2 & 6.2 \\ \hline
stereovision2 & 18.0 & 17.1 & 17.1 & 17.5 & 17.6 & 17.7 \\ \hline
Geometric mean & 23.7 & 23.5 & 23.6 & 23.9 & 23.6 & 23.7 \\ \hline
\end{tabular}
\caption{Critical path delay (in \textit{ns}) when \textit{number of cuts = 3} and \textit{increased delay = 0ns}.}
\label{table:delay0}
\end{table}

\begin{table}[!htbp]
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\% wires cut & 0 & 20 & 40 & 60 & 70 & 80 \\ \hline \hline
bgm & 31.6 & 30.3 & 29.2 & 29.2 & 29.9 & 31.6 \\ \hline
LU32PEEng & 119 & 114 & 115 & 114 & 115 & 115 \\ \hline
LU8PEEng & 118 & 116 & 119 & 117 & 114 & 115 \\ \hline
mcml & 82.6 & 82.7 & 79.6 & 81.9 & 84.5 & 82.6 \\ \hline
mkDelayWorker32B & 9.8 & 10.6 & 10.4 & 9.6 & 10.2 & 9.1 \\ \hline
stereovision0 & 5.7 & 5.3 & 4.8 & 4.9 & 4.9 & 4.8 \\ \hline
stereovision1 & 8.0 & 7.5 & 7.4 & 8.2 & 7.9 & 8.0 \\ \hline
stereovision2 & 20.1 & 19.7 & 20.3 & 19.9 & 20.2 & 21.6 \\ \hline
Geometric mean & 27.5 & 26.9 & 26.4 & 26.5 & 26.7 & 26.8 \\ \hline
\end{tabular}
\caption{Critical path delay (in \textit{ns}) when \textit{number of cuts = 3} and \textit{increased delay = 0.5ns}.}
\label{table:delay500}
\end{table}

\begin{table}[!htbp]
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\% wires cut & 0 & 20 & 40 & 60 & 70 & 80 \\ \hline \hline
bgm & 47.2 & 52.4 & 43.1 & 42.7 & 44.8 & 48.3 \\ \hline
LU32PEEng & 115 & 112 & 117 & 114 & 116 & 113 \\ \hline
LU8PEEng & 161 & 159 & 122 & 146 & 118 & 129 \\ \hline
mcml & 94.0 & 86.0 & 93.9 & 88.7 & 77.3 & 87.5 \\ \hline
mkDelayWorker32B & 15.3 & 13.8 & 14.3 & 15.3 & 17.9 & 15.4 \\ \hline
stereovision0 & 7.5 & 9.4 & 8.9 & 7.8 & 9.6 & 8.3 \\ \hline
stereovision1 & 14.6 & 12.0 & 13.1 & 12.7 & 14.6 & 14.8 \\ \hline
stereovision2 & 29.9 & 30.3 & 28.7 & 30.3 & 30.1 & 29.8 \\ \hline
Geometric mean & 37.7 & 37.3 & 35.9 & 36.1 & 37.0 & 36.9 \\ \hline
\end{tabular}
\caption{Critical path delay (in \textit{ns}) when \textit{number of cuts = 3} and \textit{increased delay = 1.5ns}.}
\label{table:delay1500}
\end{table}

@@

Tables \ref{table:1cut_minW}, \ref{table:1cut_path}, \ref{table:2cut_minW}, \ref{table:2cut_path}, \ref{table:standard_minW} and \ref{table:standard_path} show the minimum channel width and critical path delay for each circuit for each of the different scenarios. The critical path results for individual circuits again follow the overall averages well. Some of the minimum channel widths for individual circuits show anomalous trends as the number of cutlines increase, however. For example, \textit{mcml} often has a smaller minimum channel width with 2 cuts than with 1 cut. This may be due to a circuit structure that leads to easier division into thirds than halves, but also likely indicates room for further CAD optimization.

@@

\begin{table}[!htbp]
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\% wires cut & 0 & 20 & 40 & 60 & 70 & 80 \\ \hline \hline
bgm & 114 & 114 & 122 & 126 & 142 & 156 \\ \hline
LU32PEEng & 170 & 176 & 182 & 206 & 318 & 396 \\ \hline
LU8PEEng & 110 & 114 & 116 & 138 & 164 & 236 \\ \hline
mcml & 108 & 114 & 106 & 124 & 150 & 238 \\ \hline
mkDelayWorker32B & 74 & 78 & 80 & 88 & 98 & 138 \\ \hline
stereovision0 & 62 & 60 & 62 & 76 & 132 & 146 \\ \hline
stereovision1 & 102 & 104 & 106 & 108 & 144 & 186 \\ \hline
stereovision2 & 160 & 158 & 164 & 200 & 264 & 432 \\ \hline
Geometric mean & 107 & 109 & 111 & 126 & 165 & 221 \\ \hline
\end{tabular}
\caption{Minimum channel width when \textit{number of cuts = 1} and \textit{increased delay = 1ns}.}
\label{table:1cut_minW}
\end{table}

>>>>

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are left untouched, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

@@

%   - Interposer-Routing interface
\subsection{Interposer-Routing interface}
\label{sec:interposer-routing-interface}
In a simple architecture, the vertical routing wires that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

\begin{enumerate}[label=(\alph*)]
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. When \textit{fanin transfer} option is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect some adjacent ``dangling" vertical routing wires to the multiplexer. This allows for additional flexibility in routing by ensuring that \textit{every} vertical routing wire could \textit{potentially} drive an interposer wire through a multiplexer. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 multiplexer at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

\item Fan-out Transfer for Interposer Wires\\
When \textit{fanout transfer} option is \textit{on}, we allow the interposer wire output to drive not only its own corresponding vertical wire at the other side of the cut, but also the other ``dangling" vertical routing wires that are not driven by anything because their corresponding interposer wires are cut.

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{CAD Enhancements}
\label{cadSection}

\subsection{Placer optimization}
\label{cad_enh_placer_subsection}
Placement is crucial in mitigating the impact of the reduced wiring and increased delay when crossing between dice; a good placement should minimize the number of signals crossing between dice, particularly time-critical ones. VPR uses 2 different costs as the metrics for its placer algorithm: the timing cost and the bounding box (wiring) cost. We modified the placer's bounding box and timing costs to account for the increased latency to cross the interposer and for the reduced wire capacity close to the cutlines, as the wire availability becomes more sparse.

@@

\subsection{Router optimization}
\label{sec:router_optimizations}
Once the Routing Resource Graph is modified as detailed in Section \ref{archSection}, the VPR router adapted automatically to the interposer architecture. The router starts at a net source node and expands by discovering adjacent routing nodes until it reaches the destination node. At any point, the router knows the cost of the path taken to the current point and it uses a \textit{lookahead cost function} to estimate the resource and delay cost from the current node to the destination. Using this lookahead cost, the router can properly sort routing alternatives. The lookahead cost function in VPR is not aware of the interposer, and therefore it takes a long time to converge to a routing decision. To make the lookahead aware of the interposer, we modify the lookahead cost function by adding terms that account for extra \textit{interposer delay} and extra \textit{interposer nodes} that may exist on the route from the current node to the destination node.

@@

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the ``flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}). The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

We used VPR 7.0, with the enhancements we detailed above, and the architecture file \textit{k6\_frac\_N10\_mem32K\_40nm} in the experiments below. All experiments targeted the smallest FPGA (with number of rows equal to number of columns) which could accommodate a benchmark circuit; this represents a very full FPGA with little white space left, and hence presents a difficult case to an interposer-based FPGA as no die can be left mostly empty.

We have used the eight largest circuits from the VTR benchmark~\cite{vtr2012}, namely: \textit{bgm}, \textit{LU8PEEng}, \textit{LU32PEEng}, \textit{mcml}, \textit{mkDelayWorker32B}, \textit{stereovision0}, \textit{stereovision1} and \textit{stereovision2}. The size of these circuits ranges from $9,100$ to $153,000$ primitives (LUTs, FFs, etc.), with an average size of $52,600$ primitives. All results are the geometric mean over all circuits for a given interposer architecture.

To obtain the \textit{critical path delay} the circuits were run with a low stress routing with a channel width, $W = 1.3 \times minW$, where $minW$ is the minimum channel width for which the circuit is still routable.

\subsubsection{Effectiveness of placer optimizations}
\label{sec:effectiveness_of_placer_enhancements}
Enabling the enhancements described in Section~\ref{cad_enh_placer_subsection} improves the results .... haven't finished running these yet. So, <TODO EHSAN>: fill this section with actual results. For now, let's use a graph from Andre as a placeholder!
The graphs show that combining the timing with the bounding box optimizations led to an average of 20\% improvement on the area-delay product and the most significant contribution was made by incorporation of \textit{cut\_cost} into the wire cost function.

We believe the key to the good performance of the new placer cost functions is that it produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later smaller placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalize bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{bb.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

@@

\subsubsection{Effectiveness of router optimizations}
\label{sec:effectiveness_of_router_enhancements}
The router optimizations described in Section~\ref{sec:router_optimizations} extensively improve the runtime of the router lookahead algorithm. Figure~\ref{fig:lookahead_runtime} shows geomean of the router runtime for the 8 circuits for different \textit{\% wires cut} values. On average, the router runs 31 times faster and finds identical minimum channel widths for all circuits.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt.eps}
\caption{Router runtime improvement using updated lookahead function}
\label{fig:lookahead_runtime}
\end{figure}

The new router lookahead changes the priority of the routes discovered by the router. Using the new router lookahead function increases the critical path delay by 1.3\% on average as shown in Figure~\ref{fig:lookahead_fmax}.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt_fmax.eps}
\caption{The new router lookahead increases the critical path delay.}
\label{fig:lookahead_fmax}
\end{figure}

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture results}
\label{resultsSection}

Using the best CAD settings found in Section~\ref{cadSection} and the best CAD flow described in Section~\ref{sec:JaveedSection}, we analyze the impact of the key architecture parameters: \textit{\% wires cut}, \textit{delay increase}, \textit{number of cuts}, and \textit{interposer-routing interface}. We run all the experiments using the setup explained in Section~\ref{sec:CADeffect}.

@@

@@

To analyze the impact of the number of cut wires we ran experiments varying only this parameter while leaving the other parameters constant. We use \textit{number of cuts} $= 1$ (corresponding to two dice), \textit{increased delay} $= 1ns$, \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.

Figure~\ref{fig:standard_minW} shows the graph of minimum channel width versus \textit{\% wires cut}. It can be noted from this graph that the minimum channel width is almost steady up to 60\% of wires cut, indicating that the placement engine is able to avoid saturating the interposer routing until that point. When more than 60\% of the wires are cut however (i.e. the interposer provides less than 40\% of the usual within-die routing capacity), the minimum channel width grows rapidly indicating that the interposer routing bandwidth has become a limiting factor.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{minChW.eps}
\caption{Minimum channel width vs. \textit{\% wires cut} for 2 dice and $1ns$ of \textit{delay increase}.}
\label{fig:standard_minW}
\end{figure}

The critical path delay, depicted in Figure~\ref{fig:standard_crit}, on the other hand, is not strongly influenced by the percentage of wires cut, as the critical path delay at 80\% of the wires cut is essentially the same as at 0\% wires cut. Note, however, that this same critical path delay is achieved at a much higher channel width.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{standard_crit_path_new.eps}
\caption{Critical path delay vs. \textit{\% wires cut} for 2 dice and $1ns$ of \textit{delay increase}.}
\label{fig:standard_crit}
\end{figure}

Figure \ref{fig:crossingwires} provides an alternative way to visualize the relationship between interposer routing supply and routability. This figure shows how the geometric average minimum channel width required within the FPGA dice varies as the geometric average of the \emph{absolute} number of wires crossing the interposer in each channel varies, again for 2-die system. When 110 tracks cross the interposer, the interposer channels have the same capacity as the vertical routing channels within each FPGA die. As fewer wires cross the interposer, the channel width required within the FPGA dice increases to compensate for the routing difficulty in crossing the interposer. The increase is gradual as the interposer routing is reduced from 110 tracks per channel to 34 tracks per channel; over this range the routing per channel required in the FPGA dice increases from 110 tracks per channel to 114 tracks per channel. As the routing crossing the interposer is further reduced however, it becomes very difficult to increase the within-die routing sufficiently to compensate. At 18 tracks crossing the interposer channels, for example, the within-die routing must have a channel width of \emph{188} tracks to successfully route the designs. Clearly the CAD tools have the ability to trade-off interposer routing for within-die routing over a reasonable range but below a certain level (40\% of the original within-die minimum channel width in our experiments) routability becomes almost solely limited by the wiring crossing the interposer and further reduction in interposer routing is not productive.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{numberofcrossingwires_new.eps}
\caption{Geometric mean of required intra-die minimum channel width vs. geometric mean of the number of wires crossing the interposer for 2 dice and $1ns$ of \textit{delay increase}.}
\label{fig:crossingwires}
\end{figure}

To investigate the impact of the interposer delay (\textit{delay increase}), we keep all parameters constant and we sweep the \textit{delay increase} from 0 to 1.5 \textit{ns}. We use \textit{number of cuts} $= 1$ (corresponding to two dice), \textit{bidirectional interposer wires = on}, \textit{fanin transfer = on}, and \textit{fanout transfer = on}.
